Informing policy via dynamic models: Eliminating cholera in Haiti

### 2022-09-15 ms.Rnw
HAITIPKG UPDATE, moved estimation of Model 1 parameters to ms

<<Setup, include=FALSE,echo=FALSE,results='hide'>>=
library(knitr)
library(pomp)
library(panelPomp)
library(spatPomp)
library(doParallel)
library(doRNG)
library(haitipkg)
library(tidyverse)

@@

# 40 cores for doob, 8 for ito
doob_cores <- 40
gl_cores <- 36

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \citep{lee20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models; this autonomy reduced the possible effect that assumptions about the dynamic system may have on the final conclusion of the study.
Despite this autonomy, and largely adhering to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20}, the consensus across the four models was that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 without additional vaccination efforts \citep{ferguson22}.

@@

@@

# load('model1/output/trendResultsJoint.rda')  # Parameters and evaluations
load('model1/output/recal_mcap_results.rda')

p1 <- best_params
rm(best_params)

gc()

# Fit model 2. Note that this takes the same ammount of time at all run
# levels.
h2_fit <- bake(
  file = paste0("model2/", rl_dir, "model2_fit.rds"), {
    fit_haiti2()
  },
  timing = FALSE
)

@@

\begin{figure}[ht]
\centering
<<Beta_trend_Figure, fig.height=2.5, fig.width=3.7, fig.align='center'>>=
ggplot() +
  geom_point(data = maxresults, aes(x = betat, y = logLik)) +
  geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
  # geom_line(data = mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
  geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$ci[2], linetype = 'dashed') +
  # geom_vline(xintercept = mcap_results$quadratic_max, col = 'red') +
  geom_vline(xintercept = mcap_results$mle, col = 'blue') +
  labs(x = "Linear Trend in Transmission", y = 'Log Likelihood') +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10))
@
\caption{\label{fig:betat}
Monte Carlo adjusted profile of $\transmissionTrend$.
The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the confidence interval.
}
\end{figure}

@@

h3_panel_sims <- bake(
  file = paste0('model3/', rl_dir, 'panelPompSims.rds'), {
    foreach(dep = departements, .combine = rbind) %dopar% {
      SIRB <- unitobjects(h3)[[dep]]
      shared <- h3@shared
      specific <- h3@specific[, dep]

@@

      sims$dep <- dep
      sims
    }
  }
)

@@

h3Spat_quants <- bake(
  file = paste0('model3/', rl_dir, 'SpatPompSims.rds'), {
    h3Spat <- haiti3_spatPomp()
    coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)

@@

    h3Spat_sims %>%
      rename(dep = unitname) %>%
      group_by(dep, time) %>%
      summarise(
        q05 = quantile(cases, 0.025, na.rm = T),
        mean = mean(cases, na.rm = T),
        q50 = quantile(cases, 0.5, na.rm = T),
        q95 = quantile(cases, 0.975, na.rm = T)
      ) %>%
      ungroup() %>%
      mutate(
        date = lubridate::date_decimal(time)
      ) %>%
      filter(date >= yearsToDate(h3Spat@t0)) %>%
      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
  },
  timing = FALSE
)

@@

# Load pre-calculated model 1 likelihood
mod1_ll <- h1_ll['ll']  # Comes from model1/output/recal_mcap_results.rda

@@

# Calculate likelihood via block-particle filter
mod3_spatPomp_res <- bake(
  file = paste0('model3/', rl_dir, 'SpatPomp_res.rds'), {

@@

    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
    results
  },
  timing = FALSE
)

@@

lee3_results <- bake(
  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {

@@

    lee3_results
  },
  timing = FALSE
)

@@

# Simulate vaccination scenario from filtering distribution. The code
# below creates a model, samples from the filtering distribution, and then
# projects the vaccination scenario, under the specified model.
h1_S0_sims <- bake(
  file = paste0("model1/", rl_dir, "S0_sims.rds"), {

@@

    # project vacc scenario, with filtering dist, and model.
    project_from_filter2(mod = s0, PF = s0_pf, nsims = h1_scen_sims)
  },
  timing = FALSE
)

@@

# Simulate vaccination scenario from filtering distribution. The code
# below creates a model, samples from the filtering distribution, and then
# projects the vaccination scenario, under the specified model.
h1_S2_sims <- bake(
  file = paste0('model1/', rl_dir, 'S2_sims.rds'), {

@@

    # project vacc scenario, with filtering dist, and model.
    project_from_filter2(mod = s2, PF = s2_pf, nsims = h1_scen_NP)
  },
  timing = FALSE
)

@@

# Simulate vaccination scenario from filtering distribution. The code
# below creates a model, samples from the filtering distribution, and then
# projects the vaccination scenario, under the specified model.
h1_S4_sims <- bake(
  file = paste0('model1/', rl_dir, 'S4_sims.rds'), {

@@

    # project vacc scenario, with filtering dist, and model.
    project_from_filter2(mod = s4, PF = s4_pf, nsims = h1_scen_NP)
  },
  timing = FALSE
)

@@

# Simulate vaccination scenario from filtering distribution. The code
# below creates a model, samples from the filtering distribution, and then
# projects the vaccination scenario, under the specified model.
h1_S3_sims <- bake(
  file = paste0('model1/', rl_dir, 'S3_sims.rds'), {

@@

    # project vacc scenario, with filtering dist, and model.
    project_from_filter2(mod = s3, PF = s3_pf, nsims = h1_scen_NP)
  },
  timing = FALSE
)

@@

# Simulate vaccination scenario from filtering distribution. The code
# below creates a model, samples from the filtering distribution, and then
# projects the vaccination scenario, under the specified model.
h1_S1_sims <- bake(
  file = paste0('model1/', rl_dir, 'S1_sims.rds'), {

@@

    # project vacc scenario, with filtering dist, and model.
    project_from_filter2(mod = s1, PF = s1_pf, nsims = h1_scen_NP)
  },
  timing = FALSE
)

@@

h3_bpf <- bake(
  file = paste0("model3/", rl_dir, "h3_bpf.rds"),
  expr = {
    bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
  },
  timing = FALSE
)

mod3_V0_res <- bake(
  file = paste0("model3/", rl_dir, "mod3_V0_res"), {

@@

  }
)

mod3_V1_res <- bake(
  file = paste0("model3/", rl_dir, "mod3_V1_res"), {

@@

  }
)

mod3_V2_res <- bake(
  file = paste0("model3/", rl_dir, "mod3_V2_res"), {

@@

  }
)

mod3_V3_res <- bake(
  file = paste0("model3/", rl_dir, "mod3_V3_res"), {

@@

  }
)

mod3_V4_res <- bake(
  file = paste0("model3/", rl_dir, "mod3_V4_res"), {

@@

  }
)
@

>>>>

<<Setup, include=FALSE,echo=FALSE,results='hide'>>=
library(knitr)
library(pomp)
library(panelPomp)
library(spatPomp)
library(foreach)
library(doParallel)
library(doRNG)
library(haitipkg)
library(tidyverse)

@@

# 40 cores for doob, 8 for ito
# doob_cores <- 40
# gl_cores <- 36

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \citep{lee20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models; this autonomy reduced the possible effect that assumptions about the dynamic system may have on the final conclusion of the study.
Despite this autonomy, and largely adhering to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20}, the consensus across the four models was that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 \citep{trevisin22}, despite the lack of a concentrated vaccination effort.

@@

<<FitModel1, echo=FALSE, message=FALSE>>=
set.seed(636813)

NP_H1        <- switch(RUN_LEVEL, 100, 1000,   2000)
NMIF_H1      <- switch(RUN_LEVEL,  10,   50,    200)
NUM_TREND    <- switch(RUN_LEVEL,   5,   10,     90)
NPROF        <- switch(RUN_LEVEL,   4,    4,     24)
NP_EVAL      <- switch(RUN_LEVEL, 200, 1000,   5000)
NREPS_EVAL   <- switch(RUN_LEVEL, cores, 20,  cores)

if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/haiti1_fit.rds")) {
  h1_MIF <- readRDS("model1/run_level_3/haiti1_fit.rds")
} else {
  h1_MIF <- bake(
    file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
      fit_haiti1(
        NP = NP_H1, 
        NMIF = NMIF_H1, 
        NUM_TREND = NUM_TREND, 
        NPROF = NPROF,
        NREPS_EVAL = NREPS_EVAL, 
        NP_EVAL = NP_EVAL, 
        ncores = cores
      )
    }
  )
}

h1_prof_res <- h1_MIF %>%
  group_by(betat) %>%
  summarize(logLik = max(ll))

mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)

# Get the best set of parameters from MIF2 search. 
p1 <- h1_MIF %>%
  filter(ll == max(ll)) %>%
  select(-ll, -ll.se) %>%
  unlist()

mod1_ll <- h1_MIF %>%
  filter(ll == max(ll)) %>%
  select(ll) %>%
  unlist()

rm(NP_H1, NMIF_H1, NUM_TREND, NPROF, NREPS_EVAL, NP_EVAL, h1_MIF)
gc()

@@

# Fit model 2. Note that this takes the same amount of time at all run
# levels.
h2_fit <- bake(
  file = paste0("model2/", rl_dir, "model2_fit.rds"), {
    fit_haiti2()
  },
  timing = FALSE
)

@@

\begin{figure}[ht]
\centering
<<Beta_trend_Figure, fig.height=2.5, fig.width=3.7, fig.align='center'>>=
ggplot() +
  geom_point(data = h1_prof_res, aes(x = betat, y = logLik)) +
  geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
  # geom_line(data = mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
  geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$ci[2], linetype = 'dashed') +
  # geom_vline(xintercept = mcap_results$quadratic_max, col = 'red') +
  geom_vline(xintercept = mcap_results$mle, col = 'blue') +
  labs(x = "Linear Trend in Transmission", y = 'Log Likelihood') +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10))
@
\caption{\label{fig:betat}
Monte Carlo adjusted profile of $\transmissionTrend$.
The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the confidence interval.
}
\end{figure}

@@

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/panelPompSims.rds")) {
  h3_panel_sims <- readRDS("model3/run_level_3/panelPompSims.rds")
} else {
  h3_panel_sims <- bake(
    file = paste0('model3/', rl_dir, 'panelPompSims.rds'), {
      foreach(dep = departements, .combine = rbind) %dopar% {
        SIRB <- unitobjects(h3)[[dep]]
        shared <- h3@shared
        specific <- h3@specific[, dep]

@@

        sims$dep <- dep
        sims
      }
    }
  )
}

@@

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/SpatPompSims.rds")) {
  h3Spat_quants <- readRDS("model3/run_level_3/SpatPompSims.rds")
} else {
  h3Spat_quants <- bake(
    file = paste0('model3/', rl_dir, 'SpatPompSims.rds'), {
      h3Spat <- haiti3_spatPomp()
      coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)

@@

      h3Spat_sims %>%
        rename(dep = unitname) %>%
        group_by(dep, time) %>%
        summarise(
          q05 = quantile(cases, 0.025, na.rm = T),
          mean = mean(cases, na.rm = T),
          q50 = quantile(cases, 0.5, na.rm = T),
          q95 = quantile(cases, 0.975, na.rm = T)
        ) %>%
        ungroup() %>%
        mutate(
          date = lubridate::date_decimal(time)
        ) %>%
        filter(date >= yearsToDate(h3Spat@t0)) %>%
        mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
    },
    timing = FALSE
  )
}

@@

@@

# Calculate likelihood via block-particle filter
if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/SpatPomp_res.rds")) {
  mod3_spatPomp_res <- readRDS("model3/run_level_3/SpatPomp_res.rds")
} else{
  mod3_spatPomp_res <- bake(
    file = paste0('model3/', rl_dir, 'SpatPomp_res.rds'), {

@@

      results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
      results
    },
    timing = FALSE
  )
}

@@

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/lee3_results.rds")) {
  lee3_results <- readRDS("model3/run_level_3/lee3_results.rds")
} else {
  lee3_results <- bake(
    file = paste0('model3/', rl_dir, 'lee3_results.rds'), {

@@

      lee3_results
    },
    timing = FALSE
  )
}

@@

# Simulate vaccination scenario from filtering distribution. The code
# below creates a model, samples from the filtering distribution, and then
# projects the vaccination scenario, under the specified model.
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S0_sims.rds")) {
  h1_S0_sims <- readRDS("model1/run_level_3/S0_sims.rds")
} else {
  h1_S0_sims <- bake(
    file = paste0("model1/", rl_dir, "S0_sims.rds"), {

@@

      # project vacc scenario, with filtering dist, and model.
      project_from_filter2(mod = s0, PF = s0_pf, nsims = h1_scen_sims)
    },
    timing = FALSE
  )
}

@@

# Simulate vaccination scenario from filtering distribution. The code
# below creates a model, samples from the filtering distribution, and then
# projects the vaccination scenario, under the specified model.
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S2_sims.rds")) {
  h1_S2_sims <- readRDS("model1/run_level_3/S2_sims.rds")
} else {
  h1_S2_sims <- bake(
    file = paste0('model1/', rl_dir, 'S2_sims.rds'), {

@@

      # project vacc scenario, with filtering dist, and model.
      project_from_filter2(mod = s2, PF = s2_pf, nsims = h1_scen_NP)
    },
    timing = FALSE
  )
}

@@

# Simulate vaccination scenario from filtering distribution. The code
# below creates a model, samples from the filtering distribution, and then
# projects the vaccination scenario, under the specified model.
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S4_sims.rds")) {
  h1_S4_sims <- readRDS("model1/run_level_3/S4_sims.rds")
} else {
  h1_S4_sims <- bake(
    file = paste0('model1/', rl_dir, 'S4_sims.rds'), {

@@

      # project vacc scenario, with filtering dist, and model.
      project_from_filter2(mod = s4, PF = s4_pf, nsims = h1_scen_NP)
    },
    timing = FALSE
  )
}

@@

# Simulate vaccination scenario from filtering distribution. The code
# below creates a model, samples from the filtering distribution, and then
# projects the vaccination scenario, under the specified model.
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S3_sims.rds")) {
  h1_S3_sims <- readRDS("model1/run_level_3/S3_sims.rds")
} else {
  h1_S3_sims <- bake(
    file = paste0('model1/', rl_dir, 'S3_sims.rds'), {

@@

      # project vacc scenario, with filtering dist, and model.
      project_from_filter2(mod = s3, PF = s3_pf, nsims = h1_scen_NP)
    },
    timing = FALSE
  )
}

@@

# Simulate vaccination scenario from filtering distribution. The code
# below creates a model, samples from the filtering distribution, and then
# projects the vaccination scenario, under the specified model.
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S1_sims.rds")) {
  h1_S1_sims <- readRDS("model1/run_level_3/S1_sims.rds")
} else {
  h1_S1_sims <- bake(
    file = paste0('model1/', rl_dir, 'S1_sims.rds'), {

@@

      # project vacc scenario, with filtering dist, and model.
      project_from_filter2(mod = s1, PF = s1_pf, nsims = h1_scen_NP)
    },
    timing = FALSE
  )
}

@@

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/h3_bpf.rds")) {
  h3_bpf <- readRDS("model3/run_level_3/h3_bpf.rds")
} else {
  h3_bpf <- bake(
    file = paste0("model3/", rl_dir, "h3_bpf.rds"),
    expr = {
      bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
    },
    timing = FALSE
  )
}

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V0_res.rds")) {
  mod3_V0_res <- readRDS("model3/run_level_3/mod3_V0_res.rds")
} else {
  mod3_V0_res <- bake(
    file = paste0("model3/", rl_dir, "mod3_V0_res.rds"), {

@@

    }
  )
}

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V1_res.rds")) {
  mod3_V1_res <- readRDS("model3/run_level_3/mod3_V1_res.rds")
} else {
  mod3_V1_res <- bake(
    file = paste0("model3/", rl_dir, "mod3_V1_res.rds"), {

@@

    }
  )
}

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V2_res.rds")) {
  mod3_V2_res <- readRDS("model3/run_level_3/mod3_V2_res.rds")
} else {
  mod3_V2_res <- bake(
    file = paste0("model3/", rl_dir, "mod3_V2_res.rds"), {

@@

    }
  )
}

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V3_res.rds")) {
  mod3_V3_res <- readRDS("model3/run_level_3/mod3_V3_res.rds")
} else {
  mod3_V3_res <- bake(
    file = paste0("model3/", rl_dir, "mod3_V3_res.rds"), {

@@

    }
  )
}

if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V4_res.rds")) {
  mod3_V4_res <- readRDS("model3/run_level_3/mod3_V4_res.rds")
} else {
  mod3_V4_res <- bake(
    file = paste0("model3/", rl_dir, "mod3_V4_res.rds"), {

@@

    }
  )
}
@
